---
title: "Mulivariate Viz"
---

Use this file for practice with the **mulivariate viz** in-class activity.  Refer to the class website for details.


```{r}
library(tidyverse)

# Import data
weather <- read.csv("https://mac-stat.github.io/data/weather_3_locations.csv") |> 
  mutate(date = as.Date(date))  

# Check out the first 6 rows
# What are the units of observation?
# Dates
head(weather)

# How many data points do we have? 
# 2367
tail(weather)

# What type of variables do we have?
# Categorical and quantitative variables are in this dataset
class("location")
class("mintemp")
class("windspeed9am")
unique(weather$location)
```


## Example 1

```{r}
ggplot(weather, aes(x = temp3pm)) +
  geom_histogram()
```


## Example 2

How do afternoon temperatures (temp3pm) differ by location?
```{r}
# Plot 1 (no facets & starting from a density plot of temp3pm)
ggplot(weather, aes(x = temp3pm, fill = location)) + 
  geom_density()
```

```{r}
# Plot 2 (no facets or densities)
ggplot(weather, aes(x = location, y = temp3pm)) +
  geom_boxplot()
```

```{r}
# Plot 3 (facets)
ggplot(weather, aes(x = location, y = temp3pm)) +
  geom_boxplot() +
  facet_wrap(~ location)
```


## Example 3

```{r}
# Don't worry about the syntax (we'll learn it soon)
woll <- weather |>
  filter(location == "Wollongong") |> 
  mutate(date = as.Date(date))  
```

```{r}
# How often does it rain today? 
# Not often
# Fill your geometric layer with the color blue.
ggplot(woll, aes(x = raintoday)) +
  geom_bar(fill = "blue")
```

```{r}
# If it does raintoday, what does this tell us about raintomorrow?
# If it does raintoday, there is an equal chance of raintomorrow. If it does not raintoday, there is a bigger chance it will not raintomorrow than it would
# Use your intuition first
ggplot(woll, aes(x = raintoday, fill = raintomorrow)) + 
  geom_bar()
```

```{r}
# Now compare different approaches

# Default: stacked bars
ggplot(woll, aes(x = raintoday, fill = raintomorrow)) + 
  geom_bar()
```

```{r}
# Side-by-side bars
ggplot(woll, aes(x = raintoday, fill = raintomorrow)) + 
  geom_bar(position = "dodge")
```

```{r}
# Proportional bars
# position = "fill" refers to filling the frame, nothing to do with the color-related fill
ggplot(woll, aes(x = raintoday, fill = raintomorrow)) + 
  geom_bar(position = "fill")
```


## Example 4

Construct a plot that illustrates how 3pm temperatures (temp3pm) vary by date in Wollongong. Represent each day on the plot and use a curve/line to help highlight the trends.
```{r}
# THINK: What variable goes on the y-axis?
# For the curve, try adding span = 0.5 to tweak the curvature
ggplot(woll, aes(x = date, y = temp3pm)) +
  geom_point() +
  geom_smooth(span = .5)
```

```{r}
# Instead of a curve that captures the general TREND,
# draw a line that illustrates the movement of RAW temperatures from day to day
# NOTE: We haven't learned this geom yet! Guess.
ggplot(woll, aes(y = temp3pm, x = date)) 
```

New Stuff

- How does the relationship between raintoday and raintomorrow vary by location?

- How does the behavior of temp3pm over date vary by location?


## Example 5 

```{r}
# Plot temp3pm vs temp9am
# Change the code in order to indicate the location to which each data point corresponds
ggplot(weather, aes(y = temp3pm, x = temp9am)) + 
  geom_text(aes(label = location))
```

```{r}
# Change the code in order to indicate the location to which each data point corresponds
# AND identify the days on which it rained / didn't raintoday
ggplot(weather, aes(y = temp3pm, x = temp9am, color = raintoday)) + 
  geom_text(aes(label = location))
```

```{r}
# How many ways can you think to make that plot of temp3pm vs temp9am with info about location and rain?
# Play around!
ggplot(weather, aes(y = temp3pm, x = temp9am, color = location)) + 
  geom_text(aes(label = raintoday))
```


## Exercise 6

```{r}
# Change the code in order to construct a line plot of temp3pm vs date for each separate location (no points!)
# wtf
ggplot(weather, aes(y = temp3pm, x = date, fill = location)) + 
  geom_point() +
  geom_smooth(span = .5)
```


## Example 7

```{r}
# Plot the relationship of raintomorrow & raintoday
# Change the code in order to indicate this relationship by location
ggplot(weather, aes(x = raintoday, fill = raintomorrow)) + 
  geom_bar(position = "fill") +
  facet_wrap(~ location)
```


## Exercise 0

```{r}
# Import and check out data
education <- read.csv("https://mac-stat.github.io/data/sat.csv")
head(education)
```


## Exercise 1

```{r}
ggplot(education, aes(x = sat)) +
  geom_histogram()
```


## Exercise 2 

Part a
```{r}
# Construct a plot of sat vs expend
# Include a "best fit linear regression model" (HINT: method = "lm")
ggplot(education, aes(x=sat, y=expend))+
  geom_point() +
  geom_smooth(method = "lm")
```
```{r}
# Construct a plot of sat vs salary
# Include a "best fit linear regression model" (HINT: method = "lm")
ggplot(education, aes(x=sat, y=salary))+
  geom_point() +
  geom_smooth(method = "lm")
```

Part b
It is surprising that with both expenditure and salary, SAT scores increase when they decrease.


## Exercise 3

```{r}
ggplot(education, aes(x=sat, y=expend, col = salary))+
  geom_point() +
  geom_smooth(method = "lm")
```


## Exercise 4

```{r}
ggplot(education, aes(y = sat, x = salary, color = cut(expend, 3))) + 
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm")
```


## Exercise 5

Part a
```{r}
ggplot(education, aes(x = fracCat)) +
  geom_bar()
```

Part b
```{r}
ggplot(education, aes(x = fracCat, y= sat)) +
  geom_boxplot()
```

```{r}
ggplot(education, aes(x= sat)) +
  geom_histogram(col = "white", binwidth = 50) +
  facet_wrap(~ fracCat)
```

Part c
```{r}
ggplot(education, aes(x= sat, y = expend, col = fracCat)) +
  geom_point()
# It seems that states where less students take the SAT spend way more on each student and also have lower SAT score than the states thats that have higher scores, lower spending, and with more students taking the SAT. 
```

Part d
It seemed that SAT scores increased when less money was spent on students. However, when the plot is split into three by the percentage of students that actually took the SAT, we see that lower SAT scores were correlated with higher spending in states that had fewer students taking the SAT. States that had more students taking the SAT seems to show no relationship between expenditure and SAT scores.


## Exercise 5.4

